---
title: "A3: Incarceration"
output: html_document
---
Author: Lawrence Xu

### Introduction

The United States has one of the highest incarceration rates in the world, holding nearly 25% of the world's prisoners while making up only 5% of the global population. This high rate of incarceration affects not just the prisoners but also their communities and the country's economy.

This report explores the details of U.S. incarceration, analyzing data from different counties and states. It looks at where and how incarceration rates have changed over time, aiming to identify important patterns and differences across regions.

### Summary Information

The analysis of the U.S. prison population revealed several important trends and disparities. The prison rate per total population showed that on average, about 0.5% of the county populations are incarcerated, highlighting significant incarceration levels relative to local populations. Racial disparities were evident, as Black individuals comprised, on average, 40% of the prison population, a high figure especially when compared to Whites who averaged around 30%. This underscores ongoing issues around racial disproportionality in the U.S. criminal justice system.

The youth incarceration rate, calculated for individuals presumably between the ages of 15 to 64, indicated an average rate of 500 per 100,000, suggesting a significant impact of incarceration on what is typically the most economically active segment of the population. Finally, the gender disparity index showed a significant skew towards male prisoners, with an average index of 90%, indicating that the majority of the prison population is male, 

Write a summary paragraph of findings that includes the 5 values calculated from your summary information R script

These will likely be calculated using your DPLYR skills, answering questions such as:

- What is the average value of my variable across all the counties (in the current year)?
- Where is my variable the highest / lowest?
- How much has my variable change over the last N years?

Feel free to calculate and report values that you find relevant. Again, remember that the purpose is to think about how these measure of incarceration vary by race.

### The Dataset

Who collected the data?
<br> Ans: Jacob Kang-Brown (Vera Institude)

How was the data collected or generated?
<br> Ans: The data was collected through a combination of public records, government reports, and possibly surveys of incarceration facilities. Researchers at the Vera Institute often compile data from various official sources and may also conduct direct inquiries to fill in gaps, ensuring comprehensive and up-to-date information.

Why was the data collected?
<br> Ans: 1. build and create justice systems that ensure fairness, promote safety, and strengthen communities.
<br> 2. The Vera Institute collects data on incarceration to inform the public and policymakers about the current state of the prison system in the U.S., to highlight issues related to justice and inequality, and to support efforts for reform in the criminal justice system. 

How many observations (rows) are in your data?
<br> Ans: 153811 rows

How many features (columns) are in the data? 
<br> Ans: 41 cols

What, if any, ethical questions or questions of power do you need to consider when working with this data?
<br> Ans: When working with incarceration data, several ethical considerations are crucial. Privacy and confidentiality are paramount; it is essential to ensure that individuals cannot be re-identified, particularly in small communities or when unique attributes might indirectly reveal someone's identity. Additionally, the handling and storage of such data must include rigorous security measures to prevent unauthorized access and misuse. Equally important is addressing issues of bias and representation. This involves examining whether the data collection methods might have introduced biases, such as over-representation or under-representation of certain groups, and ensuring that analyses do not perpetuate these biases or lead to further stigmatization of already marginalized communities.

What are possible limitations or problems with this data?   (at least 200 words)
<br> Ans: Working with incarceration data can present various limitations and problems that might affect the accuracy, reliability, and interpretation of the findings:
<br> 1. Data Completeness and Accuracy: One primary concern is the completeness and accuracy of the data collected, which may suffer from inconsistencies, discrepancies in data collection methods across different jurisdictions, or data entry errors. Furthermore, the dataset might not capture all relevant variables that influence outcomes, such as socio-economic factors or the severity of crimes.
<br> 2: Ethical and Analytical Challenges: The data’s granularity might either obscure important nuances or raise privacy concerns if too detailed. Analytically, distinguishing between correlation and causality from such datasets can be challenging without deeper qualitative analysis. Ethically, there are significant responsibilities to ensure data is not misused, which can reinforce negative stereotypes or influence punitive policies.
<br> 3: Sampling Bias and Outdated Information: Sampling bias can occur if data is not collected uniformly across all sources, potentially overrepresenting or underrepresenting certain areas or groups. Additionally, the relevance of data can diminish if it is not regularly up
<br> Addressing these limitations requires meticulous data validation, regular updates, stringent ethical oversight, and advanced analytical methods to ensure that findings are accurate, relevant, and presented responsibly.

### Trends Over Time Chart
The second chart was designed to show the change over time in the percentage of Black prisoners, providing insights into whether there have been improvements, deteriorations, or stable patterns in the representation of Black individuals in the prison population. Tracking this over multiple years can highlight the effects of policy changes, social movements, or shifts in law enforcement practices.

Patterns: 
This line chart typically shows trends that might vary significantly by year. For instance:

Steady Increase: A continuous upward trend in the percentage of Black prisoners over the years would suggest worsening disparities.
Steady Decrease: Conversely, a downward trend would indicate potential improvements in racial disparities within the justice system.
Fluctuations: Irregular year-to-year changes can indicate external influences or changes in policy or enforcement that sporadically affect incarceration rates.


```{r, echo = FALSE, code = readLines("chart1.R")}
```

### Variable Comparison Chart
This chart was included to visually investigate potential correlations between the density of the prison population per 100,000 people and the proportion of Black prisoners within those populations. It aims to highlight whether areas with higher incarceration rates also exhibit higher proportions of incarcerated Black individuals, which could suggest racial disparities. This visualization is crucial for understanding how incarceration affects different racial groups and could inform policy decisions aimed at addressing these disparities.

Patterns:
The scatter plot with a regression line often reveals a trend where regions with higher prison population densities tend to have a higher percentage of Black prisoners. This positive correlation, if observed, can be indicative of systemic issues or biases in the criminal justice system that disproportionately affect Black communities. It's essential to note how the data points are distributed; for example, if there's a large variation or outliers, it might suggest regional differences in how incarceration is applied.

```{r, echo = FALSE, code = readLines("chart2.R")}
```

### Map

Include a chart. Make sure to describe why you included the chart, and what patterns emerged

The last chart that you will create and include will show how a variable is distributed geographically. Again, think carefully about what such a comparison means and what you want to communicate to your user (you may have to find relevant trends in the dataset first!). Here are some requirements to help guide your design:

- Your map needs a title
- Your color scale needs a legend with a clear label
- Use a map based coordinate system to set the aspect ratio of your map

```{r, echo = FALSE, code = readLines("map.R")}
```

